Probabilistic Forward Modeling of Galaxy Catalogs with Normalizing Flows

### 2023-11-28 src/tex/ms.tex
More text revisions.

% Define document class
\documentclass[twocolumn,twocolappendix,linenumbers]{aastex631}
\usepackage{showyourwork}

@@

Normalizing flows are a generative model of this variety.
Normalizing flows model complex, high-dimensional probability distributions using deep neural networks that learn an invertible mapping between the complicated data distribution and a more simple distribution, known as the latent distribution.
This allows you to sample from and calculate probabilities with respect to the latent distribution, and use the normalizing flow to translate these values back to the space of the original data distribution.
A common choice for the latent distribution is the Normal distribution, hence the name \emph{normalizing} flow.
Unlike other deep generative models like Generative Adversarial Networks (GANs; \citealt{goodfellow2014}) and Variational Autoencoders (VAEs; \citealt{kingma2014}), normalizing flows provide a deterministic mapping between the data and latent spaces.
It is this feature of normalizing flows that enables us to provide an exact answer to the question ``under the generative model, what is the posterior distribution for redshift given the simulated photometry''.
This allows us to define a ``true'' redshift posterior for each galaxy in our simulated catalog, making normalizing flows a very powerful tool for validating photo-z inference.

@@

In Section \ref{sec:nf} we provide the background on normalizing flows required to understand PZFlow, which we describe in Section \ref{sec:pzflow}.
In Section \ref{sec:fwd-model}, we demonstrate using PZFlow to simulate a galaxy catalog where each object has a redshift, photometry in the six LSST bands, a true photo-z posterior, a size, and an ellipticity.
In Section \ref{sec:photo-z}, we demonstrate using PZFlow as a density estimator by estimating photo-z's for our simulated catalog.
We conclude in Section \ref{sec:conclusion}.

@@

Figure \ref{fig:two-moons} shows an example of a normalizing flow that transforms the scikit-learn \citep{sklearn} two moons distribution into a uniform distribution.
The data points are colored by quadrant to visualize their image under $f$.

The following sections discuss how to build a normalizing flow to model data with various features.
Section \ref{sec:bijections} discusses the bijection $f$ and introduces the building blocks from which our bijections will be built;
Section \ref{sec:latent} discusses how to choose an appropriate latent distribution for your data;
Section \ref{sec:conditional} describes how to build a flow that models a conditional distribution;
Section \ref{sec:periodic} explains how to model data with periodic topology;
finally Section \ref{sec:discrete} explains how to model data with discrete variables.

@@

RQ-RSCs are bijections that are composed of coupling layers \citep{dinh2015, dinh2017}.
A coupling layer partitions the data, $x \in \R^D$, into two sets, $x_{1:d}$ and $x_{d+1:D}$.
The first set is then used to transform the second set:
\begin{align}
    \begin{split}
    y^{}_{1:d} &= x^{}_{1:d} \\
    y^{}_{d+1:D} &= g(x^{}_{d+1:D}; m(x^{}_{1:d})),
    \end{split}
\end{align}
where $g : \R^{D-d} \times \R^d \to \R^{D-d}$ is an invertible \emph{coupling law}, and $m$ is a \emph{coupling function} defined on $\R^d$.
This is illustrated in Figure \ref{fig:coupling}.
The advantage of this structure is that the Jacobian is triangular,
\begin{align}
    \frac{\partial y}{\partial x} =
    \begin{pmatrix}
         I_d & 0 \\
         \frac{\partial y^{}_{d+1:D}}{\partial x^{}_{1:d}}
         & \frac{\partial y^{}_{d+1:D}}{\partial x^{}_{d+1:D}}
    \end{pmatrix},
\end{align}
where $I_d$ is the $d \times d$ identity matrix.
In particular, the Jacobian determinant is
\begin{align}
    \det \frac{\partial y}{\partial x} = \det \frac{\partial y^{}_{d+1:D}}{\partial x^{}_{d+1:D}}.
\end{align}
Furthermore, the inverse can be calculated as
\begin{align}
    \begin{split}
    x^{}_{1:d} &= y^{}_{1:d} \\
    x^{}_{d+1:D} &= g^{-1}(y^{}_{d+1:D}; m(x^{}_{1:d})),
    \end{split}
\end{align}
Notice that neither inverting a coupling layer $g$, nor calculating the Jacobian determinant, requires inverting or taking derivatives of the coupling function $m$, which can thus be arbitrarily complex.

@@

The result is a bijection that achieves state-of-the-art performance and efficiency for forward modeling and density estimation \citep{kobyzev2020}, and are flexible enough to model complex distributions with multiple discontinuities and hundreds of modes.
In addition, they are easily adaptable for flows with periodic topology (Section \ref{sec:periodic}).
For more details, see \citet{durkan2019}.

@@

For an example of building an application-specific data processing bijection, see the \emph{Color Transform} bijection defined in Section \ref{sec:fwd-model}, which maps galaxy magnitudes to galaxy colors.
See section \ref{sec:discrete} for data processing bijections that enable modeling of discrete data.

@@

The bijections and latent distributions discussed above can be easily adapted to directly learn conditional probability distributions:
you only need to make the replacement $f(x) \to f(x;z)$, where $z$ is a vector of conditions \citep{winkler2019}.
This is illustrated in Figure \ref{fig:conditional-coupling}, which is a modification of Figure \ref{fig:coupling} to include the input of conditional variables to the coupling function $m$.
In practice, since $m$ is usually a neural network, this amounts to just appending the conditions $z$ to the inputs of the neural network.

@@

If we make the phase change $\phi \to \phi - \pi$ so that our angles $\phi \in [-\pi, \pi]$, a RQ-NSC with $B=\pi$ automatically fulfills all four constraints.
In fact, regular RQ-NSC's impose the further condition
\begin{align}
    \nabla f(-\pi) = \nabla f(\pi) = 1 \label{eq:df=1}
\end{align}
to match an identity transform for inputs outside of the range $[-\pi, \pi]$.
By choosing a latent distribution with compact support in the range $[-\pi, \pi]$, we ensure that no samples will lie outside the range of the splines, and so we can relax the boundary condition of Equation \ref{eq:df=1} in favor of the boundary condition in Equation \ref{eq:df=df}.
Spline transforms with this relaxed boundary condition are named \emph{Circular Splines} by \citet{rezende2020}.

The circular spline construction above is easily generalized to n-spheres and n-tori: $S^n \to [-1, 1]^{n-1} \times S^1$ and $T^n \to (S^1)^n$ (see \citealt{rezende2020} for more details).
We can model the joint distribution of periodic and non-periodic variables with RQ-RSCs simply by choosing appropriate bounds $B$ for each dimension, and by swapping boundary condition \ref{eq:df=1} for condition \ref{eq:df=df} for any periodic dimensions.

@@

The rest of this papaper will focus on forward modeling a photometric galaxy catalog, and photo-z inference.
Section \ref{sec:galaxy-catalog} uses PZFlow to forward model a galaxy catalog, including photometry, spectroscopic redshifts (spec-z's), \emph{true} photo-z posteriors, ellipticities, and sizes.
Section \ref{sec:photo-z} uses PZFlow for photo-z estimation, demonstrating the power of PZFlow as a density estimator, including numerous useful features for photo-z estimation.

@@

In this section, we use PZFlow to forward model a photometric galaxy catalog for the Vera Rubin Observatory's Legacy Survey of Space and Time (LSST; \citealt{ivezic2019}).
The advantage of using a catalog generated from a normalizing flow is that we have direct access to the probability distribution from which the data is drawn, enabling us to calculate true values for derived statistical products, such as the \emph{true} photo-z redshift posterior for each galaxy.

In Section \ref{sec:fwd-model} we construct a normalizing flow to model the galaxy redshifts and photometry and generate a new simulated catalog.
In Section \ref{sec:true-posteriors}, we calculate true redshift posteriors for the new catalog.
In Section \ref{sec:fwd-model-conditional} we build a conditional flow to add additional galaxy properties to the catalog.

@@

\begin{figure*}[t]
    \script{forward_model/plot_smooth_color_distribution.py}
    \begin{centering}
        \includegraphics{figures/smooth_color_distribution.pdf}
        \caption{
            Comparing the $r-i$ vs redshift distribution for galaxy samples from CosmoDC2 without photometric noise (left), CosmoDC2 with photometric noise (middle), and from the normalizing flow (right).
            High-redshift galaxies in CosmoDC2 lie along discrete tracks in redshift-color space.
            Adding photometric noise ameliorates but does not totally remove the problem.
            PZFlow produces a catalog with a smooth redshift-color distribution.
        }
        \label{fig:smooth-color-dist}
    \end{centering}
\end{figure*}

To create a generative model of galaxy redshifts and photometry, we use the true redshifts and $ugrizy$ magnitudes from the CosmoDC2 simulation \citep{dc2, cosmodc2} of the LSST Dark Energy Science Collaboration (DESC).
We add photometric errors to the true $ugrizy$ magnitudes using the 10-year-depth LSST extended-source error model of our PhotErr package (see Appendix \ref{app:error-model}), and selected galaxies with a signal-to-noise ratio (SNR) greater than 5 in the $i$ band.
Of these, we randomly selected $10^6$ galaxies and split them into training and test sets consisting of 80\% and 20\% of the galaxies, respectively.

@@

The next layer, Shift Bounds, is the data processing bijection defined in Section \ref{sec:data-processing}, which maps the range of the data onto the support of the RQ-RSC.
Note that since Shift Bounds is on the ``other side'' of the Color Transform, we need to map the ranges of the colors $u-g$, $g-r$, etc. onto the support of the splines, instead of the original magnitudes.

The final layer is an RQ-RSC, described in detail in Section \ref{sec:rq-rsc}.
This layer performs the heavy lifting of transforming the data distribution into the uniform latent distribution.
We use $D=7$ layers to transform all 7 dimensions of our data, and set $B=5$ to match the support of the latent distribution.
We use the coupling function (a feedforward neural network with two hidden layers of 128 neurons) described in \citet{durkan2019}.
We use $K=16$ spline knots.

After training the flow (see Appendix \ref{app:training-details}), we assess the results by drawing $10^4$ galaxies from the trained flow, and plotting their distribution against  $10^4$ galaxies from the test set (Figure \ref{fig:main-corner}).
We see the normalizing flow has done an excellent job of reproducing the distribution of galaxies in CosmoDC2, without any unusual artifacts or outliers.
In addition, Figure \ref{fig:smooth-color-dist} compares the distribution of galaxy $r-i$ vs redshift.
The CosmoDC2 simulation is known to exhibit discrete tracks in this space at high redshift, due to the way galaxies were assigned to a discrete set of SED templates during simulation.
These tracks are easily visible in the left panel.
Adding photometric errors largely smooths the distribution, but close inspection reveals that discrete tracks are still present in the data, especially at $z > 1$.
The right panel shows the distribution produced by PZFlow, which has a smooth color distribution, even at high redshift.
We note that these results were obtained without any extensive hyperparameter search, and that very similar (slightly worse results) are obtained without the \texttt{ColorTransform} bijection, demonstrating the flexibility of the method to adapt to unseen data sets.

With this normalizing flow, we have an efficient, probabilistic CosmoDC2 emulator that models a smooth color-redshift distribution up to redshift 3.
We generate a catalog by sampling $10^4$ galaxies from the flow, each with noisy photometry and a true redshift.
Importantly, since we have access to the probability distribution from which the galaxies were generated, we can calculate \emph{true} redshift posteriors for each galaxy.
This is the subject of the next section.

@@

When calculating these posteriors, for each galaxy we can also marginalize over the magnitudes in any missing bands, $n$:
\begin{align}
    p(z|\mathbf{\hat{m}}) = \frac{1}{p(\mathbf{\hat{m}})} \int p(z, n, \mathbf{\hat{m}}) dn,
\end{align}
which can be calculated by evaluating $p(z, n, \mathbf{\hat{m}})$ on a grid of $z$ and possible values of $n$, summing over $n$ to yield $p(z, \mathbf{\hat{m}})$, and normalizing with respect to redshift to yield $p(z|\mathbf{\hat{m}})$.
You may wish to marginalize over all values of $n$ if the galaxy was not observed in that band.
This may occur, for example, when simulated a joint Euclid-LSST catalog \citep{euclid}, as not all galaxies will have photometry from both.
You may also wish to marginalize over all values beyond a limiting magnitude to simulate a galaxy that was observed but not detected in the corresponding band.
This might occur, for example, in the low wavelength bands of Lyman-dropout galaxies observed by LSST.

Note that this marginalization is only approximate, is weakens our ability to refer to these as ``true'' redshift posteriors.
However, increasing the resolution of the grid used for $u$ band marginalization causes the resultant posteriors to converge.
Thus, we believe it is still appropriate to treat these marginalized posteriors as the truth for the purpose of photo-z validation.

Redshift posteriors for an example galaxy are displayed in Figure \ref{fig:posteriors}.
The black posterior is calculated using the full set of galaxy magnitudes.
The true redshift, marked with by the vertical red line, nearly coincides with the mode of this posterior.
The blue posterior has been calculated while marginalizing over the $u$ band.
Throwing away the information in the $u$ band slightly broadens the posterior and shifts it toward higher redshifts.

Calculating these posteriors enables direct comparison of true redshift posteriors (consistent with photometric errors and missing bands) with the redshift posteriors estimated by photo-z estimators.
This is important, as modern cosmology analyses are beginning to increasingly rely on full redshift posteriors \citep{mandelbaum2008,newman2022}.
\citePZt showed that popular metrics for evaluating photo-z estimators using ensembles of photo-z posteriors can be misleading, and are not well suited to the needs of precision cosmology.
PZFlow catalogs with true redshift posteriors provide a path forward by enabling the evaluation of photo-z estimators on a per-posterior basis.

@@

% this figure is so high just to move it ahead one page in the document
\begin{figure*}[t]
    \script{photo-z/plot_ensemble_posteriors.py}
    \begin{centering}
        \includegraphics{figures/ensemble_posteriors.pdf}
        \caption{
            The ensemble of posteriors for three example galaxies.
            Flows 1-4 label the individual posteriors produced by each of the flows that make up the ensemble.
            The dashed black line is the mean of these individual posteriors and is the value used by the ensemble.
            The vertical gray line labeled ``Truth'' denotes the true redshift of the galaxy.
            Averaging the posteriors from each flow in the ensemble approximately marginalizes over the neural network parameters, and smooths over the small-scale variations found in the posterior from any individual flow.
        }
        \label{fig:ensemble-posteriors}
    \end{centering}
\end{figure*}

In addition to the galaxy magnitudes and redshifts modeled above, we wish to include other galaxy properties in the catalog, such as galaxy size and ellipticity.
In principle, we could have included these variables in the original normalizing flow.
However, we did not want the true redshift posteriors to be conditioned on these variables, as most photo-z estimators only use galaxy photometry.
Therefore, we will build a second flow that models these additional values conditioned on the galaxy redshift and magnitudes.
In other words, we are modeling the full joint distribution via the decomposition
\begin{align}
    p(z, \mathbf{\hat{m}}, s, e) = p(z, \mathbf{\hat{m}}) \cdot p(s, e | z, \mathbf{\hat{m}}),
    \label{eq:chain-rule}
\end{align}
where $s$ is the size (the half-light radius in arcseconds) and $e$ is the ellipticity.
The first distribution on the right hand side of Equation \ref{eq:chain-rule} is modeled by our original flow, and the second distribution will be modeled using the new conditional flow.
Note that while we have only chosen to model these additional two properties, any other values you desire can be similarly modeled.

For the latent distribution, we again use a Uniform distribution over the range $[-5, 5]$.
For the bijection, we use
\begin{align}
    \begin{split}
        f =& ~ \text{RQ-RSC} \circ \text{Shift Bounds}. \\
    \end{split}
\end{align}
The RQ-RSC acts on the two dimensional space of size and ellipticity, but also takes the galaxy redshift and magnitudes as inputs (see the conditional variables in green in Figure \ref{fig:conditional-coupling}).
The redshifts and magnitudes are transformed to have zero mean and unit variance before being input to the neural network\footnote{These variables are standard scaled instead of mapped onto the domain [-5, 5], because the neural network that parameterizes the splines has no limit on inputs, unlike the splines themselves, which are limited to the range [-5, 5].} that parameterizes the splines.
Aside from the change in inputs, the RQ-RSC has the same settings as listed for the previous normalizing flow.

After training the flow (see Appendix \ref{app:training-details}), we sample a size and ellipticity for each galaxy in the PZFlow catalog created in the previous section (conditioned on the redshift and noisy magnitudes), and plot the distribution of these features against the distribution in the test set (Figure \ref{fig:conditional-corner}).
Once again, we see the normalizing flow does a good job of emulating the CosmoDC2 galaxy distribution.

The final simulated catalog consists of $10^4$ galaxies, each with a redshift, noisy $ugrizy$ magnitudes, a true photo-z posterior, a size, and an ellipticity.
We use the magnitudes, size, and ellipticity to estimate the photometric errors using the 10-year-depth LSST extended-source error model of PhotErr.
This small catalog was generated for visualization purposes, but the normalizing flows can be used to generate catalogs of arbitrarily large size.
In particular, one can generate catalogs with a millions of galaxies, or can augment existing catalogs, in only a few minutes, substantially faster than is required by large-scale simulations like CosmoDC2.

@@

In addition to forward modeling, normalizing flows are powerful and flexible models for density estimation.
This makes them useful tools for estimating posterior distributions for galaxy properties, conditioned on observed features of the galaxy.
In this section, we apply this to photo-z estimation for the simulated catalog from the previous section.

@@

We train an ensemble of 4 normalizing flows, each with the same architecture and training schedule as the regular flow described in Section \ref{sec:galaxy-catalog}.
With PZFlow, this is as simple as swapping \texttt{FlowEnsemble} for \texttt{Flow} in the code.

>>>>

% Define document class
\documentclass[twocolumn,twocolappendix]{aastex631}
\usepackage{showyourwork}

@@

Normalizing flows are a generative model of this variety.
Normalizing flows model complex, high-dimensional probability distributions using deep neural networks that learn an invertible mapping between the complicated data distribution and a more simple distribution, known as the latent distribution.
This allows you to sample from and calculate probabilities with respect to the latent distribution, and use the normalizing flow to translate these values back to the space of the original data distribution.
A common choice for the latent distribution is the Normal distribution, hence the name \emph{normalizing} flow.

Unlike other deep generative models like Generative Adversarial Networks (GANs; \citealt{goodfellow2014}) and Variational Autoencoders (VAEs; \citealt{kingma2014}), normalizing flows provide a deterministic mapping between the data and latent spaces.
It is this feature of normalizing flows that enables us to provide an exact answer to the question ``under the generative model, what is the posterior distribution for redshift given the simulated photometry''.
This allows us to define a ``true'' redshift posterior for each galaxy in our simulated catalog, making normalizing flows a very powerful tool for validating photo-z inference.

@@

In Section~\ref{sec:nf} we provide the background on normalizing flows required to understand PZFlow, which we describe in Section~\ref{sec:pzflow}.
In Section~\ref{sec:galaxy-catalog}, we demonstrate using PZFlow to simulate a galaxy catalog where each object has a redshift, photometry in the six LSST bands, a true photo-z posterior, a size, and an ellipticity.
In Section~\ref{sec:photo-z}, we demonstrate using PZFlow as a density estimator by estimating photo-z's for our simulated catalog.
We conclude in Section~\ref{sec:conclusion}.

@@

Figure~\ref{fig:two-moons} shows an example of a normalizing flow that transforms the scikit-learn \citep{sklearn} two moons distribution into a uniform distribution.
The data points are colored by quadrant to visualize their image under $f$.

The following sections discuss how to build a normalizing flow to model data with various features.
Section~\ref{sec:bijections} discusses the bijection $f$ and introduces the building blocks from which our bijections will be built;
Section~\ref{sec:latent} discusses how to choose an appropriate latent distribution for your data;
Section~\ref{sec:conditional} describes how to build a flow that models a conditional distribution;
Section~\ref{sec:periodic} explains how to model data with periodic topology;
finally Section~\ref{sec:discrete} explains how to model data with discrete variables.

@@

RQ-RSCs are bijections that are composed of coupling layers \citep{dinh2015, dinh2017}.
A coupling layer partitions the data, $x \in \R^D$, into two sets, $x_{1:d}$ and $x_{d+1:D}$.
The first set is then used to transform the second set:
\begin{align}
    \begin{split}
    y^{}_{1:d} &= x^{}_{1:d} \\
    y^{}_{d+1:D} &= g(x^{}_{d+1:D}; m(x^{}_{1:d})),
    \end{split}
\end{align}
where $g : \R^{D-d} \times \R^d \to \R^{D-d}$ is an invertible \emph{coupling law}, and $m$ is a \emph{coupling function} defined on $\R^d$.
This is illustrated in Figure~\ref{fig:coupling}.
The advantage of this structure is that the Jacobian is triangular,
\begin{align}
    \frac{\partial y}{\partial x} =
    \begin{pmatrix}
         I_d & 0 \\
         \frac{\partial y^{}_{d+1:D}}{\partial x^{}_{1:d}}
         & \frac{\partial y^{}_{d+1:D}}{\partial x^{}_{d+1:D}}
    \end{pmatrix},
\end{align}
where $I_d$ is the $d \times d$ identity matrix.
In particular, the Jacobian determinant is
\begin{align}
    \det \frac{\partial y}{\partial x} = \det \frac{\partial y^{}_{d+1:D}}{\partial x^{}_{d+1:D}}.
\end{align}
Furthermore, the inverse can be calculated as
\begin{align}
    \begin{split}
    x^{}_{1:d} &= y^{}_{1:d} \\
    x^{}_{d+1:D} &= g^{-1}(y^{}_{d+1:D}; m(x^{}_{1:d})),
    \end{split}
\end{align}
Notice that neither inverting a coupling layer $g$, nor calculating the Jacobian determinant, requires inverting or taking derivatives of the coupling function $m$, which can thus be arbitrarily complex.

@@

The result is a bijection that achieves state-of-the-art performance and efficiency for forward modeling and density estimation \citep{kobyzev2020}, and are flexible enough to model complex distributions with multiple discontinuities and hundreds of modes.
In addition, they are easily adaptable for flows with periodic topology (Section~\ref{sec:periodic}).
For more details, see \citet{durkan2019}.

@@

For an example of building an application-specific data processing bijection, see the \emph{Color Transform} bijection defined in Section~\ref{sec:fwd-model}, which maps galaxy magnitudes to galaxy colors.
See section~\ref{sec:discrete} for data processing bijections that enable modeling of discrete data.

@@

The bijections and latent distributions discussed above can be easily adapted to directly learn conditional probability distributions:
you only need to make the replacement $f(x) \to f(x;z)$, where $z$ is a vector of conditions \citep{winkler2019}.
This is illustrated in Figure~\ref{fig:conditional-coupling}, which is a modification of Figure~\ref{fig:coupling} to include the input of conditional variables to the coupling function $m$.
In practice, since $m$ is usually a neural network, this amounts to just appending the conditions $z$ to the inputs of the neural network.

@@

If we make the phase change $\phi \to \phi - \pi$ so that our angles $\phi \in [-\pi, \pi]$, a RQ-NSC with $B=\pi$ automatically fulfills all four constraints.
In fact, regular RQ-NSC's impose the further condition
\begin{align}
    \nabla f(-\pi) = \nabla f(\pi) = 1 \label{eq:df=1}
\end{align}
to match an identity transform for inputs outside of the range $[-\pi, \pi]$.
By choosing a latent distribution with compact support in the range $[-\pi, \pi]$, we ensure that no samples will lie outside the range of the splines, and so we can relax the boundary condition of Equation~\ref{eq:df=1} in favor of the boundary condition in Equation~\ref{eq:df=df}.
Spline transforms with this relaxed boundary condition are named \emph{Circular Splines} by \citet{rezende2020}.

The circular spline construction above is easily generalized to n-spheres and n-tori: $S^n \to [-1, 1]^{n-1} \times S^1$ and $T^n \to (S^1)^n$ (see \citealt{rezende2020} for more details).
We can model the joint distribution of periodic and non-periodic variables with RQ-RSCs simply by choosing appropriate bounds $B$ for each dimension, and by swapping boundary condition~\ref{eq:df=1} for condition~\ref{eq:df=df} for any periodic dimensions.

@@

The rest of this paper will focus on forward modeling a photometric galaxy catalog, and photo-z inference.
Section~\ref{sec:galaxy-catalog} uses PZFlow to forward model a galaxy catalog, including photometry, spectroscopic redshifts (spec-z's), true photo-z posteriors, ellipticities, and sizes.
Section~\ref{sec:photo-z} uses PZFlow for photo-z estimation, demonstrating the power of PZFlow as a density estimator, including numerous useful features for photo-z estimation.

@@

In this section, we use PZFlow to forward model a photometric galaxy catalog for the Vera Rubin Observatory's Legacy Survey of Space and Time (LSST; \citealt{ivezic2019}).
The advantage of using a catalog generated from a normalizing flow is that we have direct access to the probability distribution from which the data is drawn, enabling us to calculate true values for derived statistical products, such as the true photo-z redshift posterior for each galaxy.

In Section~\ref{sec:fwd-model} we construct a normalizing flow to model the galaxy redshifts and photometry and generate a new simulated catalog.
In Section~\ref{sec:true-posteriors}, we calculate true redshift posteriors for the new catalog.
In Section~\ref{sec:fwd-model-conditional} we build a conditional flow to add additional galaxy properties to the catalog.

@@

\begin{figure*}[t]
    \script{forward_model/plot_smooth_color_distribution.py}
    \begin{centering}
        \includegraphics{figures/smooth_color_distribution.pdf}
        \caption{
            Comparing the $r-i$ vs redshift distribution for galaxy samples from CosmoDC2 without photometric noise (left), CosmoDC2 with photometric noise (middle), and from the normalizing flow (right).
            High-redshift galaxies in CosmoDC2 lie along discrete tracks in redshift-color space.
            Adding photometric noise somewhat smooths but does not totally remove these tracks.
            PZFlow produces a catalog with a smooth redshift-color distribution.
        }
        \label{fig:smooth-color-dist}
    \end{centering}
\end{figure*}

To create a generative model of galaxy redshifts and photometry, we use the true redshifts and $ugrizy$ magnitudes from the CosmoDC2 simulation \citep{dc2, cosmodc2} of the LSST Dark Energy Science Collaboration (DESC).
We add photometric errors to the true $ugrizy$ magnitudes using the 10-year-depth LSST extended-source error model of our PhotErr package (see Appendix~\ref{app:error-model}), and selected galaxies with a signal-to-noise ratio (SNR) greater than 5 in the $i$ band.
Of these, we randomly selected $10^6$ galaxies and split them into training and test sets consisting of 80\% and 20\% of the galaxies, respectively.
We then train a normalizing flow to learn the distribution $p(z, \mathbf{\hat{m}})$, where $z$ is the true redshift, and $\mathbf{\hat{m}}$ is the vector of noisy magnitudes in the LSST bands.

@@

The next layer, Shift Bounds, is the data processing bijection defined in Section~\ref{sec:data-processing}, which maps the range of the data onto the support of the RQ-RSC.
Note that since Shift Bounds is on the ``other side'' of the Color Transform, we need to map the ranges of the colors $u-g$, $g-r$, etc. onto the support of the splines, instead of the original magnitudes.

The final layer is an RQ-RSC, described in detail in Section~\ref{sec:rq-rsc}.
This layer performs the heavy lifting of transforming the data distribution into the uniform latent distribution.
We use $D=7$ layers to transform all 7 dimensions of our data, and set $B=5$ to match the support of the latent distribution.
We use the coupling function (a feedforward neural network with two hidden layers of 128 neurons) described in \citet{durkan2019}.
We use $K=16$ spline knots.
This number was chosen to be large enough to capture the complexity of the data, but small enough so that the flow smooths over the discrete tracks in Figure~\ref{fig:smooth-color-dist} (more on this below).

After training the flow (see Appendix~\ref{app:training-details}), we assess the results by drawing $10^4$ galaxies from the trained flow, and plotting their distribution against  $10^4$ galaxies from the test set (Figure~\ref{fig:main-corner}).
We see the normalizing flow has done an excellent job of reproducing the distribution of galaxies in CosmoDC2, without any unusual artifacts or outliers.
In addition, Figure~\ref{fig:smooth-color-dist} compares the distribution of galaxy $r-i$ vs redshift.
High-redshift galaxies in the CosmoDC2 simulation lie on discrete tracks in this space due to the way galaxies were assigned to a discrete set of SED templates during simulation.
These tracks are easily visible in the left panel.
Adding photometric errors somewhat smooths the distribution, but close inspection reveals there is still granularity in the distribution of high-redshift galaxies.
The right panel shows the distribution produced by PZFlow, which has a smooth color distribution, even at high redshift.
We note that these results were obtained without any extensive hyperparameter search, and that very similar (slightly worse results) are obtained without the \texttt{ColorTransform} bijection, demonstrating the flexibility of the method to adapt to unseen data sets.

With this normalizing flow, we have an efficient, probabilistic CosmoDC2 emulator that models a smooth color-redshift distribution up to redshift 3.
We generate a catalog by sampling $10^4$ galaxies from the flow, each with noisy photometry and a true redshift.
Importantly, since we have access to the probability distribution from which the galaxies were generated, we can calculate true redshift posteriors for each galaxy.
This is the subject of the next section.

@@

When calculating these posteriors, for each galaxy we can also marginalize over the magnitudes in any missing bands.
Imagine, for a galaxy, that we partition the vector of magnitudes $\mathbf{\hat{m}}$ into an observed set $\mathbf{\hat{m}}_0$ and a missing set $\mathbf{\hat{m}}_\text{x}$.
We can marginalize over the missing magnitudes when calculating the posterior
\begin{align}
    p(z|\mathbf{\hat{m}}_0) = \frac{1}{p(\mathbf{\hat{m}}_0)} \int p(z, \mathbf{\hat{m}}_0, \mathbf{\hat{m}}_\text{x}) d\mathbf{\hat{m}}_\text{x},
\end{align}
which can be calculated by evaluating $p(z, \mathbf{\hat{m}})$ on a grid of $z$ and possible values of $\mathbf{\hat{m}}_\text{x}$, summing over $\mathbf{\hat{m}}_\text{x}$ to yield $p(z, \mathbf{\hat{m}}_0)$, and normalizing with respect to redshift to yield $p(z|\mathbf{\hat{m}}_0)$.
PZFlow possess a flexible method for performing this marginalization: the grid for each band in $\mathbf{\hat{m}}_\text{x}$ can be a function of other galaxy properties (e.g. the observed magnitudes, $\mathbf{\hat{m}}_0$).

You may wish to marginalize over all values of $\mathbf{\hat{m}}_\text{x}$ if the galaxy was not observed in those bands.
This may occur, for example, when simulated a joint Euclid-LSST catalog \citep{euclid}, as not all galaxies will have photometry from both.
You may also wish to marginalize over all values beyond the limiting magnitudes to simulate a galaxy that was observed but not detected in the corresponding bands.
This might occur, for example, in the low wavelength bands of Lyman-dropout galaxies observed by LSST.

Note this marginalization is only approximate, and therefore weakens our ability to refer to these as ``true'' redshift posteriors.
However, increasing the resolution of the grid used for $u$ band marginalization causes the resultant posteriors to converge.
Thus, we believe it is still appropriate to treat these marginalized posteriors as the truth for the purpose of photo-z validation.

Redshift posteriors for an example galaxy are displayed in Figure~\ref{fig:posteriors}.
The black posterior is calculated using the full set of galaxy magnitudes.
The true redshift, marked by the vertical red line, nearly coincides with the mode of this posterior.
The blue posterior has been calculated while marginalizing over the $u$ band.
Throwing away the information in the $u$ band slightly broadens the posterior and shifts it toward higher redshifts.

Calculating these posteriors enables direct comparison of true redshift posteriors with the redshift posteriors estimated by photo-z estimators.
This is important, as modern cosmology analyses are beginning to increasingly rely on full redshift posteriors \citep{mandelbaum2008,newman2022}.
\citePZt showed that popular metrics for evaluating photo-z estimators using ensembles of photo-z posteriors can be misleading, and are not well suited to the needs of precision cosmology.
PZFlow catalogs with true redshift posteriors provide a path forward by enabling the evaluation of photo-z estimators on a per-posterior basis.

@@

% this figure is so high just to move it ahead one page in the document
\begin{figure*}[t]
    \script{photo-z/plot_ensemble_posteriors.py}
    \begin{centering}
        \includegraphics{figures/ensemble_posteriors.pdf}
        \caption{
            The ensemble of posteriors for three example galaxies.
            Flows 1-4 label the individual posteriors produced by each of the flows that make up the ensemble.
            The dashed black line is the mean of these individual posteriors and is the value used by the ensemble.
            The vertical gray line labeled ``Truth'' denotes the true redshift of the galaxy.
            Averaging the posteriors from each flow in the ensemble approximately marginalizes over the neural network parameters, and smooths over the small-scale variations found in the posterior from any individual flow.
            The first panel is a galaxy with a narrow and unimodal redshift posterior, while the next two panels demonstrate broad, multimodal posteriors, which is typical of galaxies in the range $1.5 < z < 2.6$.
        }
        \label{fig:ensemble-posteriors}
    \end{centering}
\end{figure*}

In addition to the galaxy magnitudes and redshifts modeled above, we wish to include other galaxy properties in the catalog, such as galaxy size and ellipticity.
In principle, we could have included these variables in the original normalizing flow.
However, we did not want the true redshift posteriors to be conditioned on these variables, as most photo-z estimators only use galaxy photometry.
Therefore, we will build a second flow that models these additional values conditioned on the galaxy redshift and magnitudes.
In other words, we are modeling the full joint distribution via the decomposition
\begin{align}
    p(z, \mathbf{\hat{m}}, s, e) = p(z, \mathbf{\hat{m}}) \cdot p(s, e | z, \mathbf{\hat{m}}),
    \label{eq:chain-rule}
\end{align}
where $s$ is the size (the half-light radius in arcseconds) and $e$ is the ellipticity.
The first distribution on the right hand side of Equation~\ref{eq:chain-rule} is modeled by our original flow, and the second distribution will be modeled using the new conditional flow.
While we have only chosen to model these additional two properties, any other values you desire can be similarly modeled.

For the latent distribution, we again use a Uniform distribution over the range $[-5, 5]$.
For the bijection, we use
\begin{align}
    \begin{split}
        f =& ~ \text{RQ-RSC} \circ \text{Shift Bounds}. \\
    \end{split}
\end{align}
The RQ-RSC acts on the two dimensional space of size and ellipticity, but also takes the galaxy redshift and magnitudes as inputs (see the conditional variables in green in Figure~\ref{fig:conditional-coupling}).
The redshifts and magnitudes are transformed to have zero mean and unit variance before being input to the neural network\footnote{These variables are standard scaled instead of mapped onto the domain [-5, 5], because the neural network that parameterizes the splines has no limit on inputs, unlike the splines themselves, which are limited to the range [-5, 5].} that parameterizes the splines.
Aside from the change in inputs, the RQ-RSC has the same settings as listed for the previous normalizing flow.

After training the flow (see Appendix~\ref{app:training-details}), we sample a size and ellipticity for each galaxy in the PZFlow catalog created in the previous section (conditioned on the redshift and noisy magnitudes), and plot the distribution of these features against the distribution in the test set (Figure~\ref{fig:conditional-corner}).
Once again, we see the normalizing flow does a good job of emulating the CosmoDC2 galaxy distribution.
We note that, if desired, the TARP test of \citet{lemos2023} can provide a quantitative test of the fidelity of the conditional flow.

The final simulated catalog consists of $10^4$ galaxies, each with a redshift, noisy $ugrizy$ magnitudes, a true photo-z posterior, a size, and an ellipticity.
We use the magnitudes, size, and ellipticity to estimate the photometric errors using the 10-year-depth LSST extended-source error model of PhotErr.
This small catalog was generated for visualization purposes, but the normalizing flows can be used to generate catalogs of arbitrarily large size.
In particular, in only a few minutes, one can generate new catalogs or augment existing catalogs with millions of galaxies.
This is substantially faster than re-running large scale simulations like CosmoDC2.

@@

In addition to forward modeling, normalizing flows are powerful and flexible models for density estimation.
This makes them useful tools for estimating posterior distributions for galaxy properties, conditioned on observed features of the galaxy.
In this section, we demonstrate this by applying PZFlow to photo-z estimation for the simulated catalog from the previous section.

@@

We train an ensemble of 4 normalizing flows, each with the same architecture and training schedule as the regular flow described in Section~\ref{sec:galaxy-catalog}.
With PZFlow, this is as simple as swapping \texttt{FlowEnsemble} for \texttt{Flow} in the code.
